---
title: "Spatial Connectivity"
author: "Nick Gauthier"
output:
  html_document: 
    fig_height: 10
    fig_width: 10
    keep_md: yes
---
# Data import

First import the SWSN attribute file. Use tidyverse packages for data munging.

Site coordinates are in UTM, so first use rgdal to reproject to LatLon.
```{r message = F, warning = F}
library(tidyverse)
library(sp) 
library(rgrass7)
library(pbapply)
```

Now reimport the attribute file, select the relevant data, and combine with the reprojected site coordinates.
```{r message = F}
swsn.attr <- read_csv('Data/attributes_orig.csv') %>%
  select(ID = as.numeric(SWSN_ID), site = SWSN_Site, 
         macro = Macro, micro = Micro, 
         easting = as.numeric(EASTING), northing = as.numeric(NORTHING)) %>%
  unite(coors, easting, northing, sep = ',', remove = F) ## also add column of both easting and northing
```
Use the function to import the network datasets.
```{r}
readSWSN <- function(net){
  read_csv(net) %>%
  select(site = X1) %>%
  semi_join(swsn.attr, .) %>%
  SpatialPointsDataFrame(.[,c(6,7)], ., proj4string=CRS("+proj=utm +zone=12 +datum=WGS84"))
}
```
Use the function to import the network datasets.
```{r}
ad1200 <- readSWSN('Data/AD1200sim.csv')
ad1250 <- readSWSN('Data/AD1250sim.csv')
ad1300 <- readSWSN('Data/AD1300sim.csv')
ad1350 <- readSWSN('Data/AD1350sim.csv')
ad1400 <- readSWSN('Data/AD1400sim.csv')
```

Use the rgrass7 library to try doing the same thing in GRASS

```{r}

execGRASS('g.region', res = '100')

# create an empty raster map to add the shortest paths to after each iteration
execGRASS('r.mapcalc', flags = c('overwrite'), expression = 'fete = 0')
execGRASS('r.mapcalc', flags = c('overwrite'), expression = 'friction = 1')


## calculate the shortest paths to all sites from each site in turn in a for loop



dem <- 'NED_100m'
### in the future can just get a text string for the points and feed them to r.walk, sending dem thru writeRAST
getCost <- function(sites, grass_pts){
  
  from_sites <- sites$coors %>% c
  to_sites <- sites$coors %>% c %>% paste(collapse=',') %>% strsplit(',') %>% unlist %>% as.numeric
  
  pbsapply(from_sites, function(x){
        from_site <- x %>% strsplit(',') %>% unlist %>% as.numeric
        execGRASS('r.walk', elevation = dem,
                      friction = 'friction',
                      output = 'walk_costs',
                      start_coordinates = from_site,
                      stop_coordinates = to_sites,
                      memory = 1500,
                      flags = c('overwrite', 'quiet'))
  
    execGRASS('r.what', map = 'walk_costs', points = grass_pts, separator = ',', intern = T) %>%
      data_frame(x = .) %>% 
      separate(x, c('x', 'y', 'cat','cost'), sep = ',', convert = T) %>%
      select(cost) %>%
      c %>%
      return
  }) %>% do.call(rbind,.)
}

ad1200_costs <- getCost(ad1200, 'ad1200')
write.csv(ad1200_costs, 'Data/ad1200_costs.csv', row.names = F, col.names = F)

ad1250_costs <- getCost(ad1250, 'ad1250')
write.csv(ad1250_costs, 'Data/ad1250_costs.csv', row.names = F, col.names = F)

ad1300_costs <- getCost(ad1300, 'ad1300')
write.csv(ad1300_costs, 'Data/ad1300_costs.csv', row.names = F, col.names = F)


ad1350_costs <- getCost(ad1350, 'ad1350')
ad1400_costs <- getCost(ad1400, 'ad1400')





execGRASS('r.drain',
                      input = 'walk_costs',
                      direction = 'walk_dir',
                      output = 'shortest_paths',
                      drain = 'paths_vect',
                      start_points = sites,
                      flags = c('d', 'overwrite'))


    for point in listpoints.splitlines():
        # extract a starting point from the sites map
        # and create a new point file with it
        print 'starting point extracted'
        doGRASS('v.extract',
                          input = sites,
                          where = 'cat = ' + point,
                          output = 'startpoint',
                          overwrite = 'True',
                          quiet = 'True')
        # calculate the anisotropic cost surface
        # from the previously extacted point to all the sites using knight's move
        print 'starting r.walk'
        grass.run_command('r.walk',
                      elevation = dem,
                      friction = frict,
                      output = 'walk_costs',
                      outdir = 'walk_dir',
                      start_points = 'startpoint',
                      stop_points = sites,
                      flags = 'k',
                      overwrite = 'True')
        # use the outputs of r.walk to create a raster of the shortest paths to each site
        print 'starting r.drain'
        grass.run_command('r.drain',
                      input = 'walk_costs',
                      direction = 'walk_dir',
                      output = 'shortest_paths',
                      drain = 'paths_vect',
                      start_points = sites,
                      flags = 'd',
                      overwrite = 'True')
        # r.drain creates a raster map where the all cells on the path = 1,

        grass.run_command('v.to.rast',
                          input = 'paths_vect',
                          type = 'line',
                          output = 'shortest_paths',
                          use = 'val',
                          overwrite = 'True',
                          quiet = 'True')
        # uses these values to make a cumulative map
        print 'adding shortest paths to cumulative paths'
        grass.mapcalc('fete = fete + if(isnull(shortest_paths),0,1)', overwrite='True')
        print 'point ' + point + ' of ' + str(len(listpoints.splitlines())) + ' complete'
        
    # convert the cumulative fete raster to a vector and
    # assign the number of paths through each segment to the vector's attributes.
    grass.run_command('r.to.vect',
                    input = 'fete',
                    output = 'fete',
                    type = 'line',
                    column = 'paths_thru',
                    overwrite = 'True')
    # create a spatial network by combining the cumultive fete vector with the initial points vector
    print 'converting fete to network'
    grass.run_command('v.net',
                    input = 'fete',
                    points = sites,
                    output= sites + '_net',
                    operation = 'connect',
                    threshold = '1',
                    flags = 's',
                    overwrite = 'True')
```

