---
title: "Southwest Social Networks"
author: "Nick Gauthier"
output:
  pdf_document: 
    keep_tex: yes
  html_document: 
    fig_height: 10
    fig_width: 10
    keep_md: yes
---
# Data import

First import the SWSN attribute file. Use tidyverse packages for data munging.

Site coordinates are in UTM, so first use rgdal to reproject to LatLon.
```{r message = F, warning = F}
library(tidyverse)
library(rgdal)
library(igraph)
library(ggraph)
library(maps)
library(maptools)
library(raster)
library(gdistance)
```

Import the SWSN attribute file and reproject the UTM coordinates to Lat-Lon.
```{r}
swsn.attr <- read_csv('Data/attributes_orig.csv') %>%
  select(-1) # drop ID column

swsn.pts <-  swsn.attr %>% 
  select(x = EASTING, y = NORTHING) %>%
  SpatialPoints(proj4string=CRS("+proj=utm +zone=12 +datum=WGS84")) %>%
  spTransform(CRS("+proj=longlat +datum=WGS84")) %>% 
  coordinates %>%
  data.frame
```


Now define a function to import the SWSN adjacency matrix for a given time step. This function imports the adjacency matrix, keeps only those connections with >= 75% similarity, and creates an igraph object. Then it adds attribute data from above to the graph object.
```{r message = F}
readSWSN <- function(net){
  read.csv(net, row.names = 1, check.names = F)  %>%
    as.matrix %>%
    graph_from_adjacency_matrix(mode = 'undirected', weighted = T, diag = F) %>%
    as_data_frame %>%
    graph_from_data_frame(directed = F, vertices = swsn.attr)
}
```

Use the function to import the network datasets.
```{r}
ad1200 <- readSWSN('Data/AD1200sim.csv')
```

## some network plots

```{r}
ad1200 %>% 
  delete_edges(E(ad1200)[E(ad1200)$weight < .75]) %>%
  ggraph('manual', node.positions = swsn.pts) +
  geom_edge_link(aes(alpha = weight, color = weight)) +
  geom_node_point(aes(size = P1room)) +
  scale_size_area() +
  scale_edge_colour_viridis() +
  scale_edge_alpha() +
  coord_quickmap() +
  theme_void()
```

```{r}
states <- map('state', regions = c('arizona', 'new mexico'), fill = T, plot = F)
IDs <- sapply(strsplit(states$names, ":"), function(x) x[1])
states.ply <- map2SpatialPolygons(states, IDs=IDs)
```

```{r}
swsn.layout <- ad1200 %>% 
  delete_edges(E(ad1200)[E(ad1200)$weight < .5]) %>%
  create_layout('manual', node.positions = swsn.pts) 
```

```{r}
ggraph(swsn.layout) +
  geom_polygon(aes(x = long, y = lat, group = region), data = states, color = 'black', fill = NA) +
  geom_edge_link(aes(alpha = weight, color = weight)) +
  geom_node_point(aes(size = P1room)) +
  scale_size_area() +
  #scale_edge_color_viridis() +
  scale_edge_color_distiller() +
  scale_edge_alpha() +
  geom_label(x = -106, y = 35, label = 'AD 1200') +
  coord_quickmap() +
  theme_void() +
  theme(legend.position="none")
```

now without pointsizing
```{r}
ggraph(swsn.layout) +
  geom_polygon(aes(x = long, y = lat, group = region), data = states, color = 'black', fill = NA) +
  geom_edge_link(aes(alpha = weight, color = weight)) +
  geom_node_point() +
  #scale_edge_color_viridis() +
  scale_edge_color_distiller() +
  scale_edge_alpha() +
  geom_label(x = -106, y = 35, label = 'AD 1200') +
  coord_quickmap() +
  theme_void() +
  theme(legend.position="none")
```
Now without any points
```{r}
ggraph(swsn.layout) +
  geom_polygon(aes(x = long, y = lat, group = region), data = states, color = 'black', fill = NA) +
  geom_edge_link(aes(alpha = weight, color = weight)) +
  #scale_edge_color_viridis() +
  scale_edge_color_distiller() +
  scale_edge_alpha() +
  geom_label(x = -106, y = 35, label = 'AD 1200') +
  coord_quickmap() +
  theme_void() +
  theme(legend.position="none")
```
log transform the colors
```{r}
ggraph(swsn.layout) +
  geom_polygon(aes(x = long, y = lat, group = region), data = states, color = 'black', fill = NA) +
  geom_edge_link(aes(alpha = weight, color = weight)) +
  #scale_edge_color_viridis() +
  scale_edge_color_distiller() +
  scale_edge_alpha() +
  geom_label(x = -106, y = 35, label = 'AD 1200') +
  coord_quickmap() +
  theme_void() +
  theme(legend.position="none")
```

```{r}
library(RColorBrewer)
getPalette = colorRampPalette(brewer.pal(9, "Set1"))((swsn.layout$Macro %>% unique %>% length))

ggraph(swsn.layout) +
  geom_polygon(aes(x = long, y = lat, group = region), data = states, color = 'black', fill = NA) +
  geom_bin2d(aes(x = x, y = y, fill = Macro), na.rm=TRUE) +
  scale_fill_manual(values = getPalette) +
  geom_label(x = -106, y = 35, label = 'AD 1200') +
  coord_quickmap() +
  theme_void() +
  theme(legend.position="none")
```

## Least cost distances
LCP distance matrix

```{r}
elev.agg <- raster('~/Data/NED/swsn_NED_1.tif') %>% aggregate(fact = 10)
plot(elev.agg)
```

```{r}
altDiff <- function(x){x[2] - x[1]}
hd <- transition(elev.agg, altDiff, 8, symm=FALSE)
slope <- geoCorrection(hd, type = 'c')

adj <- adjacent(elev.agg, cells=1:ncell(elev.agg), pairs=TRUE, directions=8)
speed <- slope
speed[adj] <- 6 * exp(-3.5 * abs(slope[adj] + 0.05))
Conductance <- geoCorrection(speed, type = 'c')
rm(slope, hd, adj, speed, elev.agg)


lcp <- costDistance(Conductance, as.matrix(swsn.pts)) %>%
  ./3600 %>%
  graph_from_adjacency_matrix(mode = 'directed', weighted = T, diag = F) %>%
  set_vertex_attr(name, V(ad1200$name))

swsn.lcp.layout <- lcp %>% 
  induced_subgraph(as_ids(V(delete_vertices(ad1200, degree(ad1200) == 0)))) %>%
  delete_edges(E(.)[E(.)$weight > 5]) %>%
  #set_vertex_attr('eigen.c', value = eigen_centrality(.)) %>%
  create_layout('manual', node.positions = swsn.pts[match(V(test)$name, swsn.attr$SWSN_Site),]) 
```

```{r}
ggraph(swsn.lcp.layout) +
  geom_polygon(aes(x = long, y = lat, group = region), data = states, color = 'black', fill = NA) +
  geom_label(x = -106, y = 35, label = 'AD 1200') +
  geom_edge_link(alpha = .1) +
  geom_node_point(color = 'red') +
  scale_size_area() +
  coord_quickmap() +
  theme_void()
```



Now try adding the EOF
```{r}
eof.dat <- brick('Data/eof1200.nc')[[3]] %>% 
  mask(states.ply) %>%
  as.data.frame(xy = T, na.rm = T) %>%
  mutate(value = cut(X3, breaks=seq(-max(abs(X3)), max(abs(X3)), length.out=11)))

qplot(x, y, fill = value, dat = eof.dat, geom = 'raster') +
  scale_fill_brewer(palette = 'RdBu')

ggraph(swsn.layout) +
  geom_raster(aes(fill = value), data = eof.dat) +
  geom_polygon(aes(x = long, y = lat, group = region), data = states, color = 'black', fill = NA) +
  geom_edge_link(aes(alpha = weight, color = weight)) +
  geom_node_point() +
    scale_fill_brewer(palette = 'RdBu') +
  #scale_edge_color_viridis() +
  scale_edge_color_distiller() +
  scale_edge_alpha() +
  geom_label(x = -106, y = 35, label = 'AD 1200') +
  coord_quickmap() +
  theme_void() +
  theme(legend.position="none") 



```



## More Minimal network maps

```{r}
library(GGally)
library(ggmap)
library(raster)


plotEOF <- function(x){
  mask(x, states.ply) %>%
  rasterVis::gplot +
  geom_raster(aes(fill = value), na.rm = T, show.legend = F) +
  scale_fill_distiller(palette = 'RdBu', na.value = NA) +
  geom_polygon(data = states, aes(x = long, y = lat, group = region), color = 'black', fill = NA) +
  coord_quickmap() +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude")
}

```

