---
title: "Statistical Modeling of SWSN Data"
output: html_notebook
---

Integrating network data and eofs


First import the SWSN attribute file. Use tidyverse packages for data munging.

Site coordinates are in UTM, so first use rgdal to reproject to LatLon.
```{r message = F, warning = F}
library(raster)
library(rasterVis)
library(tidyverse)
library(rgdal)

swsn.pts <- read_csv('Data/attributes_orig.csv') %>% 
  select(easting = EASTING, northing = NORTHING) %>%
  SpatialPoints(proj4string=CRS("+proj=utm +zone=12 +datum=WGS84")) %>%
  spTransform(CRS("+proj=longlat +datum=WGS84")) %>% 
  coordinates %>%
  data.frame
```

Now reimport the attribute file, select the relevant data, and combine with the reprojected site coordinates.
```{r message = F}
swsn.attr <- read_csv('Data/attributes_orig.csv') %>%
  dplyr::select(ID = SWSN_ID, site = SWSN_Site, macro = Macro, micro = Micro, size = P1room) %>%
  cbind(swsn.pts)
```

Now define a function to import the SWSN adjacency matrix for a given time step. This function imports the adjacency matrix and creates an igraph object. Then it adds attribute data from above to the graph object.
```{r message = F}
library(igraph)

readSWSN <- function(net){
  net.in <- read.csv(net, row.names = 1, check.names = F) %>% 
    as.matrix %>%
    graph_from_adjacency_matrix(mode = 'undirected', weighted = T, diag = F)
  
  ord <- match(V(net.in)$name, swsn.attr$site)

  V(net.in)$lon <- swsn.attr[ord, 6]
  V(net.in)$lat <- swsn.attr[ord, 7]
  V(net.in)$size <- swsn.attr[ord, 5]
  V(net.in)$region <- swsn.attr[ord, 3] %>% as.character
  
  return(net.in)
}
```

Use the function to import the network datasets.
```{r}
ad1200 <- readSWSN('Data/AD1200sim.csv')
ad1250 <- readSWSN('Data/AD1250sim.csv')
ad1300 <- readSWSN('Data/AD1300sim.csv')
ad1350 <- readSWSN('Data/AD1350sim.csv')
ad1400 <- readSWSN('Data/AD1400sim.csv')
```

Import eof data
```{r}
eof <- brick('Data/eof_all.nc')[[1:6]] %>% setNames(c('eof1', 'eof2', 'eof3', 'eof4', 'eof5', 'eof6'))
levelplot(eof, par.settings = RdBuTheme(), at = seq(-.03, .03, .003))
```
Import elevation too.
```{r}
elev <- raster('Data/swsn_NED_1.tif')
levelplot(elev, margin = F)
```
Also mean climate data.
```{r}
prcp.mean <- brick('Data/prec-sw.nc') %>% stackApply(.,indices = rep(1: (nlayers(.)/12), each = 12), fun = sum) %>% mean
levelplot(prcp.mean, margin = F, par.settings = PuOrTheme())
```

```{r}
temp.mean <- ((brick('Data/trefmnav-sw.nc') %>% mean) + (brick('Data/trefmnav-sw.nc') %>% mean)) / 2
levelplot(temp.mean)
```

```{r}
water.stress.mean <- brick('Data/water_stress.nc') %>% stackApply(.,indices = rep(1: (nlayers(.)/12), each = 12), fun = sum) %>% mean
levelplot(water.stress.mean)
plot(water.stress.mean, elev)
```


```{r}
ad1200.pts <- cbind(V(ad1200)$lon, V(ad1200)$lat)
eof.vals <- raster::extract(eof, ad1200.pts)
V(ad1200)$eof1 <- eof.vals[,1]
V(ad1200)$eof2 <- eof.vals[,2]
V(ad1200)$eof3 <- eof.vals[,3]
V(ad1200)$eof4 <- eof.vals[,4]
V(ad1200)$eof5 <- eof.vals[,5]
V(ad1200)$eof6 <- eof.vals[,6]

V(ad1200)$elev <- raster::extract(elev, ad1200.pts)
V(ad1200)$prcp <- raster::extract(prcp.mean, ad1200.pts)
V(ad1200)$temp <- raster::extract(temp.mean, ad1200.pts)
V(ad1200)$ws <- raster::extract(water.stress.mean, ad1200.pts)

el <- get.edgelist(ad1200)
E(ad1200)$eof1 <- abs(V(ad1200)[el[, 1]]$eof1 - V(ad1200)[el[, 2]]$eof1)
E(ad1200)$eof2 <- abs(V(ad1200)[el[, 1]]$eof2 - V(ad1200)[el[, 2]]$eof2)
E(ad1200)$eof3 <- abs(V(ad1200)[el[, 1]]$eof3 - V(ad1200)[el[, 2]]$eof3)
E(ad1200)$eof4 <- abs(V(ad1200)[el[, 1]]$eof4 - V(ad1200)[el[, 2]]$eof4)
E(ad1200)$eof5 <- abs(V(ad1200)[el[, 1]]$eof5 - V(ad1200)[el[, 2]]$eof5)
E(ad1200)$eof6 <- abs(V(ad1200)[el[, 1]]$eof6 - V(ad1200)[el[, 2]]$eof6)

# now binary
E(ad1200)$eof1.b <- sign(V(ad1200)[el[, 1]]$eof1) != sign(V(ad1200)[el[, 2]]$eof1)
E(ad1200)$eof2.b <- sign(V(ad1200)[el[, 1]]$eof2) != sign(V(ad1200)[el[, 2]]$eof2)
E(ad1200)$eof3.b <- sign(V(ad1200)[el[, 1]]$eof3) != sign(V(ad1200)[el[, 2]]$eof3)
E(ad1200)$eof4.b <- sign(V(ad1200)[el[, 1]]$eof4) != sign(V(ad1200)[el[, 2]]$eof4)
E(ad1200)$eof5.b <- sign(V(ad1200)[el[, 1]]$eof5) != sign(V(ad1200)[el[, 2]]$eof5)
E(ad1200)$eof6.b <- sign(V(ad1200)[el[, 1]]$eof6) != sign(V(ad1200)[el[, 2]]$eof6)

E(ad1200)$from_lat <- V(ad1200)[el[, 1]]$lat
E(ad1200)$from_lon <- V(ad1200)[el[, 1]]$lon
E(ad1200)$to_lat <- V(ad1200)[el[, 2]]$lat
E(ad1200)$to_lon <- V(ad1200)[el[, 2]]$lon


E(ad1200)$size.prod <- V(ad1200)[el[, 1]]$size * V(ad1200)[el[, 2]]$size


E(ad1200)$elev <- abs(V(ad1200)[el[, 1]]$elev - V(ad1200)[el[, 2]]$elev)
E(ad1200)$prcp <- abs(V(ad1200)[el[, 1]]$prcp - V(ad1200)[el[, 2]]$prcp)
E(ad1200)$temp <- abs(V(ad1200)[el[, 1]]$temp - V(ad1200)[el[, 2]]$temp)
E(ad1200)$water.stress <- abs(V(ad1200)[el[, 1]]$ws - V(ad1200)[el[, 2]]$ws)


dat <- igraph::as_data_frame(ad1200)  %>%
  dplyr::mutate_at(c('eof1.b', 'eof2.b', 'eof3.b', 'eof4.b', 'eof5.b', 'eof6.b'), factor)
dat
```



```{r}
library(geosphere)
dat$dist <- distHaversine(cbind(dat$from_lon, dat$from_lat), cbind(dat$to_lon, dat$to_lat))
```

LCP distance matrix
```{r}
library(gdistance)
```

```{r}
elev.agg <- raster('~/Data/NED/swsn_NED_1.tif') %>% aggregate(fact = 100)
```

```{r}
altDiff <- function(x){x[2] - x[1]}
hd <- transition(elev.agg, altDiff, 8, symm=FALSE)
slope <- geoCorrection(hd, type = 'c')

adj <- adjacent(elev.agg, cells=1:ncell(elev.agg), pairs=TRUE, directions=8)
speed <- slope
speed[adj] <- 6 * exp(-3.5 * abs(slope[adj] + 0.05))
Conductance <- geoCorrection(speed, type = 'c')

lcp1 <- costDistance(Conductance, cbind(dat$from_lon, dat$from_lat), cbind(dat$to_lon, dat$to_lat)) %>% diag
lcp2 <- costDistance(Conductance, cbind(dat$to_lon, dat$to_lat), cbind(dat$from_lon, dat$from_lat)) %>% diag
dat$lcp <- (lcp1 + lcp2) / 3600
```

```{r}
qplot(lcp, weight, data = dat, alpha = I(.1), geom = 'point') +
  geom_smooth() +
  theme_minimal()
```


```{r}
qplot(x = dat$dist, geom = 'density')
qplot(x = dat$lcp, geom = 'density')
qplot(dist, lcp, data = dat, geom = 'point') +
  geom_smooth() +
  theme_minimal()
```

Let's start with just the environmental covariates
```{r}
library(mgcv)
library(broom)
mod1 <- gam(weight ~ eof2.b + eof3.b +eof4.b +eof5.b +eof6.b +
              s(lcp, bs = 'cr') +
              s(ws, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof2, by = eof2.b, bs = 'cr') + 
              s(eof3, by = eof3.b, bs = 'cr') + 
              s(eof4, by = eof4.b, bs = 'cr') + 
              s(eof5, by = eof5.b, bs = 'cr') + 
              s(eof6, by = eof6.b, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', select = T, data = dat)
plot(mod1, pages = 1) 
summary(mod1)
gam.check(mod1)
glance(mod1)
tidy(mod1)
```
```{r}

mod2 <- gam(weight ~ eof3.b + eof5.b  +
              s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof2, bs = 'cr') + 
              s(eof3, by = eof3.b, bs = 'cr') + 
              s(eof4, by = eof4.b, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)
mod3 <- gam(weight ~ eof3.b +
              s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof2, bs = 'cr') + 
              s(eof3, by = eof3.b, bs = 'cr') + 
              s(eof4, by = eof4.b, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)
mod4 <- gam(weight ~ s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof3, bs = 'cr') + 
              s(eof4, by = eof4.b, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)
mod5 <- gam(weight ~ eof3.b + eof5.b  +
              s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof3, by = eof3.b, bs = 'cr') + 
              s(eof4, by = eof4.b, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)
mod6 <- gam(weight ~ s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof3, bs = 'cr') + 
              s(eof4, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)

lapply(list(mod2, mod3, mod4, mod5, mod6), glance) %>% do.call(rbind, .)

plot(mod2, pages = 1)
plot(mod3, pages = 1)
plot(mod4, pages = 1)
plot(mod5, pages = 1)
glance(mod2)
summary(mod2)
```
```{r}
EvaluateSmooths = function(model, select=NULL, x=NULL, n=100) {
  if (is.null(select)) {
    select = 1:length(model$smooth)
  }
  do.call(rbind, lapply(select, function(i) {
    smooth = model$smooth[[i]]
    data = model$model

    if (is.null(x)) {
      min = min(data[smooth$term])
      max = max(data[smooth$term])
      x = seq(min, max, length=n)
    }
    if (smooth$by == "NA") {
      by.level = "NA"
    } else {
      by.level = smooth$by.level
    }
    range = data.frame(x=x, by=by.level)
    names(range) = c(smooth$term, smooth$by)

    mat = PredictMat(smooth, range)
    par = smooth$first.para:smooth$last.para

    y = mat %*% model$coefficients[par]

    se = sqrt(rowSums(
      (mat %*% model$Vp[par, par, drop = FALSE]) * mat
    ))

    return(data.frame(
      label=smooth$label
      , x.var=smooth$term
      , x.val=x
      , by.var=smooth$by
      , by.val=by.level
      , value = y
      , se = se
    ))
  }))
}

smooths = EvaluateSmooths(mod6)
library(ggplot2)

gam.names <- c(`lcp` = "Travel time (hours)",
               `water.stress` = "Water stress (mm)",
              `elev` = "Elevation (m)",
                `eof3` = "EOF3",
              `eof4` = "EOF4"
                    )
ggplot(smooths, aes(x.val, value)) + 
  geom_line() + 
  geom_line(aes(y=value + 2*se), linetype="dashed") + 
  geom_line(aes(y=value - 2*se), linetype="dashed") + 
  facet_wrap(~ x.var, scales = 'free_x', nrow = 2, labeller = as_labeller(gam.names)) +
  labs(title = 'Predictors of tie strength', subtitle = 'Generalized additive model smooths and standard errors', x = '', y = '') +
  theme_bw() +
  theme(strip.background = element_blank())
ggsave('smooths.pdf', width = 6, height = 4.5)
```




```{r}
mod3 <- gam(weight ~ eof2.b +
              s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof3, bs = 'cr') + 
              s(eof4, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)
plot(mod3, pages = 1)
glance(mod3)
summary(mod3)
```

```{r}
mod4 <- gam(weight ~ eof2.b + eof3.b + eof4.b +
              s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof3, by = eof3.b, bs = 'cr') + 
              s(eof4, by = eof4.b, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)
mod5 <- gam(weight ~ eof3.b + eof4.b +
              s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof3, by = eof3.b, bs = 'cr') + 
              s(eof4, by = eof4.b, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)
mod6 <- gam(weight ~ eof2.b + 
              s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof3, by = eof3.b, bs = 'cr') + 
              s(eof4, by = eof4.b, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)

mod7 <- gam(weight ~ s(lcp, bs = 'cr') +
              s(water.stress, bs = 'cr') +
              s(elev, bs = 'cr') +
              s(eof3, by = eof3.b, bs = 'cr') + 
              s(eof4, bs = 'cr'),
            family=betar(link="logit"),
            method = 'REML', data = dat)
lapply(list(mod4, mod5, mod6, mod7), glance) %>% do.call(rbind, .)

plot(mod4, pages = 1)
summary(mod6)
glance(mod4)
```
     
     
     
     
